---
title: "Modeling life expectancy with lifeplus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling life expectancy with lifeplus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4
)
```

```{r setup}
library(lifeplus)
library(ggplot2)
```

## Introduction
The `lifeplus` package allows you to fit a variety of Bayesian hierarchical models to life expectancy data. The models are modular, allowing you to mix and match multiple types of model components. 

Let $c = 1, \dots, C$ index the areas (typically countries) and $t = 1, \dots, T$ index time points (typically years). Let $y_{c,t}$ be the observed life expectancy in area $c$ at time $t$. The model takes the form

$$
\mathrm{logit}\left(y_{c,t}\right) = g\left(y_{c,t-1}, \beta_c\right) + \delta_{c,t} + \epsilon_{c,t}
$$

where

- $g$ is a _transition model_ with area-specific parameters $\beta_c$.
- $\delta_{c,t}$ is a shock term, to which a _shock model_ is applied.
- $\epsilon_{c,t}$ is a residual modeled by a _data model_.

Multiple options are available for each of these models. Each model may have additional configuration options; to learn about the options that are available, you can check the documentation of each function (e.g, `?transition_model_double_logistic`). 

### Transition models

- **Double logistic**: `transition_model_double_logistic`. The transition is modeled by a parametric double logistic function (Raftery et al. 2013). 
- **Spline**: `transition_model_spline`. The transition is modeled by a B-spline function.
- **Gaussian Process**: `transition_model_gaussian_process`. The transition function is assigned a Gaussian Process prior, approximated using a basis function expansion method.

### Shock models

- **None**: `shock_model_none`. All shock terms are fixed to zero: $\delta_{c,t} = 0$.
- **Regularized horseshoe**: `shock_model_regularized_horseshoe`. Shock terms are modeled with a regularized horseshoe prior (Piironen and Vehtari 2017). 

### Data models
- **Normal**: `data_model_normal`. The residuals are modeled as $\epsilon_{c,t} \sim N\left(0, \sigma^2\right)$.
- **Outlier**:  `data_model_outlier` . Any absolute life expectancy differences larger than a user-specified threshold are ignored. The remaining observations are modeled as $\epsilon_{c,t} \sim N\left(0, \sigma^2\right)$. 

## Fitting World Population Prospect Data

### Data
Data are from the [United Nations World Population Prospects](https://population.un.org/wpp/) R package for their 2024 report, called [`wpp2024`](https://github.com/PPgp/wpp2024):
```{r wppdata}
library(wpp2024)
data(UNlocations, package = "wpp2024")
data(e0M1, package = "wpp2024")
```

First, let's find the country codes for the countries we are interested in. To start, we will fit the model to a single country: Somalia.
```{r}
country_codes <- UNlocations |>
  dplyr::filter(name == "Somalia") |>
  dplyr::pull(country_code)
```

The WPP data is in wide format. Let's convert it to long format, with a separate column for the year (called `year`) and the life expectancy (called `e0`).
```{r}
data <- e0M1 |>
  dplyr::filter(country_code %in% country_codes) |>
  tidyr::pivot_longer(cols = `1950`:`2023`, names_to = "year", values_to = "e0") |>
  dplyr::mutate(year = as.numeric(year))
```

```{r}
ggplot(data, aes(x = year, y = e0)) +
  geom_point() +
  labs(x = "Year", y = "Male life expectancy at birth (e0)", title = "Somalia")
```

### Modeling with `lifeplus`

Now we are ready to fit the first model. Let's start with a double logistic transition model, normal data model, and no shock model (all shock terms fixed to zero):
```{r model_one}
fit <- lifeplus(
  data = data, 
  y = "e0", 
  time = "year", 
  area = "name", 
  end_time = 2050, 
  transition_model = transition_model_double_logistic(hierarchical = FALSE),
  data_model = data_model_normal(), 
  shock = shock_model_none(),
  iter_warmup = 500,
  iter_sampling = 500,
  refresh = 0,
  show_messages = FALSE,
  show_exceptions = FALSE,
  parallel_chains = 4
)
```

The `plot_projections` function plots the posterior model projections:
```{r model_one_projections}
plot_projections(fit, data = FALSE)
```

The `plot_transitions` function plots the posterior of the estimated transition function:
```{r model_one_transitions}
plot_transitions(fit)
```

Let's try fitting the model again, but this time swapping in a regularized horseshoe prior for the shock term:
```{r model_shocks}
fit_shocks <- lifeplus(
  data = data, 
  y = "e0", 
  time = "year", 
  area = "name", 
  end_time = 2050, 
  transition_model = transition_model_double_logistic(hierarchical = FALSE),
  data_model = data_model_normal(), 
  shock = shock_model_regularized_horseshoe(scale_global = 0.1),
  iter_warmup = 500,
  iter_sampling = 500,
  refresh = 0,
  show_messages = FALSE,
  show_exceptions = FALSE,
  parallel_chains = 4
)
```

```{r model_two_plots}
plot_transitions(fit_shocks)
plot_projections(fit_shocks, data = FALSE)
```

We can also plot the estimated shock terms from this model:
```{r model_two_shocks}
plot_shocks(fit_shocks)
```

## References
- Riutort-Mayol, G., Bürkner, PC., Andersen, M.R. et al. (2020) [Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming](https://link.springer.com/article/10.1007/s11222-022-10167-2). Statistics and Computing 33, 17.
- Juho Piironen, Aki Vehtari. (2017) [Sparsity information and regularization in the horseshoe and other shrinkage priors](https://doi.org/10.1214/17-EJS1337SI).
  Electronic Journal of Statistics, 11(2), 5018-5051.   
- Raftery, A.E., Chunn, J.L., Gerland, P. et al. (2013) [Bayesian probabilistic projections of life expectancy for all countries](https://doi.org/10.1007/s13524-012-0193-x). 
  Demography 50, 777–801.
- Susmann, H., Alexander, M., and Alkema, L. (2022) [Temporal Models for Demographic and Global Health Outcomes in Multiple Populations: Introducing a New Framework to Review and Standardise Documentation of Model Assumptions and Facilitate Model Comparison](https://doi.org/10.1111/insr.12491).
  International Statistical Review, 90: 437–467. 
