---
title: "Getting started with lifeplus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with lifeplus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lifeplus)
library(ggplot2)
```

Data are from the United Nations World Population Prospects 2024 report's R package `wpp2024`: 
```{r wppdata}
library(wpp2024)
data(UNlocations, package = "wpp2024")
data(e0M1, package = "wpp2024")
```

First, let's find the country codes for the countries we are interested in. To start, we will fit the model to a single country: Somalia.
```{r}
country_codes <- UNlocations |>
  dplyr::filter(name == "Somalia") |>
  dplyr::pull(country_code)
```

The WPP data is in wide format. Let's convert it to long format, with a separate column for the year (called `year`) and the life expectancy (called `e0`).
```{r}
data <- e0M1 |>
  dplyr::filter(country_code %in% country_codes) |>
  tidyr::pivot_longer(cols = `1950`:`2023`, names_to = "year", values_to = "e0") |>
  dplyr::mutate(year = as.numeric(year))
```

```{r}
ggplot(data, aes(x = year, y = e0)) +
  geom_point() +
  labs(x = "Year", y = "Male life expectancy at birth (e0)", title = "Somalia")
```

Now we are ready to fit the first model. 
```{r}
fit <- lifeplus(
  data = data, 
  y = "e0", 
  time = "year", 
  area = "name", 
  end_time = 2050, 
  transition_model = transition_model_double_logistic(hierarchical = FALSE),
  data_model = data_model_normal(), 
  shock = shock_model_none(),
  iter_warmup = 500,
  iter_sampling = 500,
  parallel_chains = 4
)
```

```{r}
plot_transitions(fit)
plot_projections(fit, data = FALSE)
```

```{r}
fit_outlier <- lifeplus(
  data = data, 
  y = "e0", 
  time = "year", 
  area = "name", 
  end_time = 2050, 
  transition_model = transition_model_double_logistic(hierarchical = FALSE),
  data_model = data_model_outlier(), 
  shock = shock_model_none(),
  iter_warmup = 500,
  iter_sampling = 500,
  parallel_chains = 4
)
```

```{r}
plot_transitions(fit_outlier)
plot_projections(fit_outlier, data = FALSE)
```
